A Practical Optional Type System for Clojure

### 2012-10-21 dissertation/design-choices.tex
Intro to variable-arity
Fixed citation breaking spaces. Latest corrections of intro

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's 
for Typed Racket \cite{Tob10}. Some of the main differences are described in this chapter.

@@

\section{Heterogeneous Maps}
% keyword keys for now, generally encouraged, fast
% Some heterogeneous maps have string args, eg. from web frameworks

As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
This also serves as a prototype for formalising other aspects.
Here we give a formal syntax, operational semantics and type system using inference
rules, as is standard in the programming language community (see, eg., Pierce\cite{Pie02}).

>>>>

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's 
for Typed Racket~\cite{Tob10}. Some of the main differences are described in this chapter.

@@

\section{Heterogeneous Maps}
\label{ref:designhmap}
% keyword keys for now, generally encouraged, fast
% Some heterogeneous maps have string args, eg. from web frameworks

\begin{verbatim}
HMaps only convey positive information, not negative.
Map concatenation is not approached, see lit review, many
complications. Instead, only support assoc, dissoc, get.
\end{verbatim}

As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
This also serves as a prototype for formalising other aspects.
Here we give a formal syntax, operational semantics and type system using inference
rules, as is standard in the programming language community (see, eg., Pierce~\cite{Pie02}).

### 2012-10-21 dissertation/experiments.tex
Implementation work
Fixed citation breaking spaces. Latest corrections of intro

\section{Monads}

Monads are an interesting control structure most used in functional programming languages,
as described by Wadler\cite{Wad95}.
They are most recognisable from its inclusion in the statically-typed language Haskell\cite{Mar10},
where monads are relied on for many features including global state, file output, and exceptions.

I chose to port the Clojure Contrib library \emph{algo.monads}\cite{Hic12} to Typed Clojure. The library
provides several kinds of monads, monad transformers, and monadic functions.
Macros are used to provide pleasant syntax for consumers of the library.

>>>>

\section{Monads}
\label{sec:monads}

Monads are an interesting control structure used in functional programming languages,
as described by Wadler~\cite{Wad95}.
They are most recognisable from its inclusion in the statically-typed language Haskell~\cite{Mar10},
where monads are relied on for many features including global state, file output, and exceptions.

I chose to port the Clojure Contrib library \emph{algo.monads}~\cite{Hic12} to Typed Clojure. The library
provides several kinds of monads, monad transformers, and monadic functions.
Macros are used to provide pleasant syntax for consumers of the library.

### 2012-10-21 dissertation/implementation.tex
Fixed citation breaking spaces. Latest corrections of intro

The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
that the expected type matches the actual type of the expression.
If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
The algorithm is based on Pierce and Turner's  Local Type Inference\cite{PT00}
and the implementation is similar in shape to Typed Racket's\cite{TF08}, in that
one function with an extra ``expected type'' argument is preferred over two complementary
functions, one for checking and one for synthesis.

@@

The basic idea of occurrence typing involves keeping a separate environment of \emph{propositions}
that relate bindings to types. These propositions are then used to update the type environment.
In Typed Clojure there are several types of propositions, referred to as \emph{filters}.
They are based on the theory by Tobin-Hochstadt and Felleisen\cite{TF10}, and are ported directly
from Typed Racket.

@@

The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
for bounded variables,
and is reminiscent to Pierce and Turner's Local Type Inference\cite{PT00}.
A common entry point for inferring type variables to polymorphic function invocations
\lstinline|typed.core/infer|.

@@

Bounds consist of an upper and lower type bound, or a \emph{kind bound}.
Kind bounds are experimental following the inclusion of user definable type constructors
(motivated in section \ref{sec:monads}).
They are a stub for a more comprehensive treatment of higher-kinded operators
such as that described by Moors, Piessens, and Odersky for Scala\cite{MPO08}.
At present, a type variable can only be instantiated to a type between
its upper and lower bounds, or, if a kind bound is defined instead, 
to a kind below the kind bound.

@@

It also required changes to the polymorphic type inference, where each
reference to a type variable required a special case for a dotted type variable.
For example, the constraint-generation algorithm for Local Type Inference
features extra kinds of constraints for dotted variables.
Strickland, Tobin-Hochstadt, and Felleisen elaborate on the particular changes
required for the Typed Racket implementation\cite{STF09}.

@@

Instead, we generate new propositions using a technique 
suggested by Tobin-Hochstadt\cite{Tob12},
that follows from the occurrence typing calculus defined 
by Tobin-Hochstadt and Felleisen\cite{TF10}.
First the filter set for functions are inferred as usual.
Then the ``then-proposition'' is applied to the environment and new propositions
are added to the filter set for each variable in the type environment that changes.
The same procedure is followed for the ``else-proposition''.

>>>>

The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
that the expected type matches the actual type of the expression.
If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
The algorithm is based on Pierce and Turner's  Local Type Inference~\cite{PT00}
and the implementation is similar in shape to Typed Racket's~\cite{TF08}, in that
one function with an extra ``expected type'' argument is preferred over two complementary
functions, one for checking and one for synthesis.

@@

The basic idea of occurrence typing involves keeping a separate environment of \emph{propositions}
that relate bindings to types. These propositions are then used to update the type environment.
In Typed Clojure there are several types of propositions, referred to as \emph{filters}.
They are based on the theory by Tobin-Hochstadt and Felleisen~\cite{TF10}, and are ported directly
from Typed Racket.

@@

The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
for bounded variables,
and is reminiscent to Pierce and Turner's Local Type Inference~\cite{PT00}.
A common entry point for inferring type variables to polymorphic function invocations
\lstinline|typed.core/infer|.

@@

Bounds consist of an upper and lower type bound, or a \emph{kind bound}.
Kind bounds are experimental following the inclusion of user definable type constructors
(motivated in section \ref{sec:monads}).
They are a stub for a more comprehensive treatment of higher-kinded operators
such as that described by Moors, Piessens, and Odersky for Scala~\cite{MPO08}.
At present, a type variable can only be instantiated to a type between
its upper and lower bounds, or, if a kind bound is defined instead, 
to a kind below the kind bound.

@@

It also required changes to the polymorphic type inference, where each
reference to a type variable required a special case for a dotted type variable.
For example, the constraint-generation algorithm for Local Type Inference
features extra kinds of constraints for dotted variables.
Strickland, Tobin-Hochstadt, and Felleisen elaborate on the particular changes
required for the Typed Racket implementation~\cite{STF09}.

@@

Instead, we generate new propositions using a technique 
suggested by Tobin-Hochstadt~\cite{Tob12},
that follows from the occurrence typing calculus defined 
by Tobin-Hochstadt and Felleisen~\cite{TF10}.
First the filter set for functions are inferred as usual.
Then the ``then-proposition'' is applied to the environment and new propositions
are added to the filter set for each variable in the type environment that changes.
The same procedure is followed for the ``else-proposition''.

### 2012-10-21 dissertation/intro.tex
Fixed citation breaking spaces. Latest corrections of intro

\emph{It is practical and useful to design and implement an optional typing system 
for the Clojure programming language that allows Clojure programmers to continue 
using idioms and style found in current Clojure code.}

@@

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This idea not new, but recent attempts are noteworthy for their successful use of bidirectional checking\footnote{These 
slides by Felleisen, a practitioner in this area for over 20 years, present a history of optional type checking
for Lisp/Scheme that spans around 40 years\cite{Fell09}}.
Instead of always attempting to infer types the algorithm relies on programmer annotations
appearing in some natural places, such as giving the type of each top-level function.

@@

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's excellent host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the extensive Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. For example, ``design by contract'' is often used,
introduced by Meyer for the Eiffel language\cite{Mey92},
in which the programmer defines contracts that are enforced at runtime.
Unit testing is also a popular verification technique in dynamic languages.
Clojure adopts these approaches, providing easy syntax for defining 
Eiffel-inspired pre- and post-conditions and a library for writing unit tests.

Static type systems, however, are still desirable for certain situations.
Powerful type systems like ML's\cite{Mil97} and Haskell's\cite{Mar10} have proved particularly useful
for writing in complicated programming styles. For example,
Haskell's advanced static type system helps the programmer write correct monadic code,
especially in more complicated situations like combining monads via monad transformers.

@@

The initial motivation for implementing an optional type system
for Clojure occurred during an anecdotal account of the struggles
of a Clojure programmer. In a heroic effort, he managed to 
implement a library for conduits (or \emph{arrows}) in Clojure, 
which probably surpass monads in complexity, and is usually reserved
for languages with advanced type systems like Haskell.

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness
of the library (before noting that unit tests and runtime contracts
were insufficient for this task).
Secondly, to aid him while writing the library.

Since then, the Clojure community has shown interest in this work.
I developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community.
I have also been invited to speak on this project at the Clojure Conj
2012 conference, a major annual Clojure conference.

@@

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning\cite{Pfe08} and Reynolds\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.

@@

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
a view strongly advocated by Harper\cite{Har12}
and common in the literature on static types (for example, Pierce\cite{Pie02}).

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Tobin-Hochstadt \cite{Tob10}}

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.

@@

Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
implementation in Typed Racket for variable-arity polymorphism \cite{STF09}
that is sufficient to handle uniform and non-uniform variable-arity functions.
Typed Clojure includes an implementation of the most immediately useful parts of variable-arity
polymorphism using algorithms, nomenclature, and implementation based on this work.

>>>>

\emph{It is practical and useful to design and implement an optional typing system 
for the Clojure programming language using bidirectional checking that allows Clojure programmers to continue 
using idioms and style found in current Clojure code.}

@@

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new, but recent attempts are noteworthy for 
their broad success in matching many of the advantages of statically typed
languages, notably due to the use of \emph{bidirectional checking}.
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

@@

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's extensive host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the extensive Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

Recently a number of languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. For example, ``design by contract'' is often used,
introduced by Meyer for the Eiffel language~\cite{Mey92},
in which the programmer defines contracts that are enforced at runtime.
Unit testing is also a popular verification technique in dynamic languages.
Clojure adopts these approaches, providing easy syntax for defining 
Eiffel-inspired pre- and post-conditions and a library for writing unit tests.

Static type systems, however, are still desirable for many situations.
Powerful type systems like ML's~\cite{Mil97} and Haskell's~\cite{Mar10} have proved particularly useful
when complicated programming styles are required. For example,
Haskell's advanced static type system helps the programmer write correct monadic code
(as detailed by Wadler~\cite{Wad95})
especially in more complicated situations like combining monads via monad transformers.

@@

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer. In an apparently heroic effort, he managed to 
implement a Clojure library for conduits, an advanced form of ``pipes'',
using arrows, a generalisation of monads.
\footnote{CONDUIT REFERENCE TODO}
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such a implementation correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012.
\footnote{CITE GSOC, or FOOTNOTE}
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the

@@

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
a view strongly advocated by Harper~\cite{Har12}
and common in the literature on static types (for example, Pierce~\cite{Pie02}).

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Tobin-Hochstadt~\cite{Tob10}}

@@

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.

@@

Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
implementation in Typed Racket for variable-arity polymorphism~\cite{STF09}
that is sufficient to handle uniform and non-uniform variable-arity functions.
Typed Clojure includes an implementation of the most immediately useful parts of variable-arity
polymorphism using algorithms, nomenclature, and implementation based on this work.
